---
title: "HW1"
author: Neal Marquez
date: "January 16th, 2019"
output: pdf_document
---

```{r libs, message=FALSE, warning=FALSE, echo=FALSE}
rm(list=ls())
library(tidyverse)
library(mlmRev)
library(knitr)
```

## Homework 1  

This homework uses data from the 1975 U.S. Sustaining Effects Study of elementary education, available as the egsingle dataset in the `mlmRev` R package. This gives data on 1,721 students in 60 schools. We will take `math` (Mathematics achievement score) as the outcome variable, and `schoolid` (the code for the school the student attends) as the grouping variable. For this homework, we will use the data for year $0.5$ only.  

### Question 1  

We will first consider the classical analysis of variance model.  

#### A) Write out the model formally using sum contrasts.  

$$
y_i = \alpha_{j[i]} + \epsilon_i,~i=1, \dots,n \\
\alpha_{j[i]} = \beta_0 + \beta_{j[i]} \\
\sum_{k=1}^M \beta_k = 0 \\
\epsilon \sim \mathcal{N}(0,\sigma^2_y)
$$


#### B) Estimate the parameters of the model.  

We may estimate the parameters of this model by Constructing a design matrix $X$ which is $N \times (M)$ in dimensions where $N$ is the number of observation and $M$ is the number of groups. The matrix is populated with the following algorithm.

$$
\begin{aligned}
&\text{if } j[i] \neq M, \\
&~~~~X_{i,1} = 1 \\
&~~~~X_{i,j[i]} = 1 \\
&~~~~X_{i,k \notin \{ 1,j[i]\}} = 0 \\
&\text{else}, \\
&~~~~X_{i,1} = 1 \\
&~~~~X_{i,k \neq 1} = -1 \\
\end{aligned}
$$
This formulation will give us an intercept $\beta_0$ that represents the mean of all the groups, $M-1~\beta$ coefficients that represent the difference between a specific group and the mean of all groups. The last groups coefficient, $\beta_M$, may be calculated from $-\sum_{k=1}^{M-1} \beta_k$. It's standard error may be calculated as $\sqrt{\sum_k^{M-1} \text{Var}(\beta_k) + \sum_k^{M-1} \sum_{s \neq k}^{M-1}\text{CoVar}(\beta_k, \beta_s)}$.

#### C) Plot the model parameters and their confidence intervals in informative ways.

Below we plot the school effects for the 60 schools along with their uncertainty sorted from smallest to largest. We see that about a third of the schools have a significant school effect than the mean of the schools.  

```{r, echo=FALSE}
data(egsingle)
schoolDF <- egsingle %>%
    filter(year == .5) %>%
    group_by(schoolid) %>%
    mutate(ahat=mean(math)) %>%
    ungroup

parEst <- lm(
    math ~ schoolid, data=schoolDF, contrasts=list(schoolid = contr.sum))

# calc the stats for the last coefficient
muLast <- -sum(coefficients(parEst)[-1])
sterrLast <- sqrt(sum(vcov(parEst)[-1,-1]))


confint(parEst) %>%
    as_tibble %>%
    mutate(coef=names(parEst$coefficients)) %>%
    mutate(est=unname(parEst$coefficients)) %>%
    filter(coef != "(Intercept)") %>%
    rbind(tibble(
        `2.5 %` = muLast - 1.98*sterrLast,
        `97.5 %` = muLast + 1.98*sterrLast,
        coef = paste0("schoolid", length(unique(schoolDF$schoolid))),
        est = muLast
    )) %>%
    arrange(-est) %>%
    mutate(id=1:n()) %>%
    mutate(Significant=(0 > `97.5 %` & 0 >`2.5 %`) | (0 < `97.5 %` & 0 <`2.5 %`)) %>%
    ggplot(aes(x=id, y=est, ymin=`2.5 %`, ymax=`97.5 %`, color=Significant)) +
    coord_flip() +
    theme_classic() +
    geom_point() +
    geom_errorbar() +
    labs(x="School Effects", y="Parameter Estimates") +
    theme(
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
    geom_hline(yintercept = 0, linetype=2) +
    ggtitle("School Effects As a Difference From the Mean(w/ 95% CI)")

```

#### D. Carry out a formal statistical test for the significance of the school effect.  
Below we show the results of an F-test testing the effects of schools and we find that school has a significant effect on the outcome of math score.


```{r, echo=FALSE}
anova(parEst)
```

#### E. What conclusions do you draw from this analysis?  

Schools significantly differ from one another in their average math scores. That is to say that we can better estimate a students predicted score if we know the school from which they come from.

### Question 2  

Using the results of your analysis in question 1, answer the following.  

#### A. Calculate the within-school variance, the between-school variance and the intraclass correlation coefficient.

Recall that these values may be calculated as seen below.

$$
\hat\sigma^2_y = \frac{1}{n-J}\sum_{i=1}^n (y_i - \hat\alpha_{j[i]})^2 \\
\hat\sigma^2_\alpha = \frac{1}{J-1}\sum_{i=1}^J (\hat\alpha_j - \bar\alpha) \\
\hat\rho = \frac{\hat\sigma^2_\alpha}{\hat\sigma^2_\alpha + \hat\sigma^2_y}
$$

This leaves us with the following estimates

```{r, echo=F}
J <- length(unique(schoolDF$schoolid))
sigma2y <- 1/(nrow(schoolDF) - J)*sum((schoolDF$math - schoolDF$ahat)^2)
sigma2alpha <- schoolDF %>%
    group_by(schoolid) %>%
    summarize(alpha_hat=mean(math)) %>%
    ungroup %>%
    mutate(avg=mean(alpha_hat), ssi=(alpha_hat - avg)^2) %>%
    select(ssi) %>%
    unlist %>%
    sum %>%
    `*`(1/(J-1))
icc <- sigma2alpha / (sigma2alpha + sigma2y)


kable(tibble(
    Statistic = c(
        "Within-School Variance",
        "Between-School Variance",
        "Intraclass Correlation Coefficient"),
    Estimate = c(sigma2y, sigma2alpha, icc)
))

ignore.ysterr <- (var(schoolDF$math) / nrow(schoolDF))^.5
account.ysterr <- summary(parEst)$coefficients[1,"Std. Error"]
```

#### B. If all the schools had the same number of students (i.e. if there were 60 schools with 29 students each), what would the design effect for a two-stage sample be?

If all the groups are the same size we may calculate the design effect as  

$$
1 + (n_1-1)\rho
$$

Which for our estimate, $\hat\rho$, leaves us with a design effect of `r round(1+(29-1)*icc, 4)`.

#### C. Ignoring the within-school correlation, what is the estimate of the overall average score? What is its standard error? By how much would the standard error be underestimated by ignoring within-school correlation? Calculate an adjusted standard error for the overall mean score taking account of within-school correlation. [When answering this sub-question, ignore differences between the sizes of schools, i.e. pretend that there are 60 schools with 29 students each.]

We can calculate an estimate of the mean and SE ignoring within school correlation using the following equations.

$$
\bar y = n^{-1}\sum_{i=1}^n y_i \\
\bar\sigma^\star_y = \text{Var}[y]n^{-1}
$$

This gives us an overall mean of `r round(mean(schoolDF$math), 4)` which, if all schools have the same number of students, should match the intercept of a sum contrasts model specification. On the other hand the estimate of the standard error for this mean is `r round(ignore.ysterr, 4)` which when compared to the estimate which accounts for in school correlation, `r round(account.ysterr, 4)`, leads us to an overestimation of the standard error of about `r round(account.ysterr/ignore.ysterr, 4)`.
