---
title: "HW8"
author: Neal Marquez
date: "March 10th, 2019"
output:
  pdf_document:
    fig_caption: yes
---

As in previous homeworks, this question uses data from the 1975 U.S. Sustaining Effects Study of elementary education, available as the `egsingle` dataset in the `mlmRev` `R` package.This gives data on 1,721 students in 60 schools. We will take `math`(Mathematics achievement score) as the outcome variable, and `schoolid`(the code for the school the student attends) as the grouping variable.  We will use only the data for year 0.5 (for which there are data on 1672 students).

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    fig.width=4.8, fig.height=3.6, echo=FALSE, warning=FALSE, message=FALSE)
```

```{r}
library(mlmRev)
library(tidyverse)
library(brms)
library(tidybayes)
library(lme4)
set.seed(123)

DF <- egsingle %>%
    filter(year == .5) %>%
    as_tibble
```

## Question 1

_Estimate the Bayesian random effects one-way analysis of variance model for these data via Markov chain Monte Carlo using the prior distribution you specified in Homework 7._

```{r include=F}
schoolDF <- DF %>%
    group_by(schoolid) %>%
    summarize(mmath=mean(math))

s20 <- var(DF$math)
mu0 <- mean(schoolDF$mmath)
t20 <- var(schoolDF$mmath)

# Check out what the names of priors can be
# get_prior(math ~ 1 + (1|schoolid), data=DF)
# check out doccumentation on possible priors
# https://mc-stan.org/docs/2_18/functions-reference

prior <- c(
    prior(normal(-0.3138749, 1.203647), class = Intercept),
    prior(inv_gamma(.5, 0.6018236), class = sigma),
    prior(inv_gamma(.5, 0.2877828), class = sd))

modelFit <- brm(
    math ~ 1 + (1|schoolid), data=DF, prior = prior,
    cores=4, iter=10000)
```

## Question 2
_Assess the convergence of your algorithm and whether it has run for enough iterations. If not, run it for longer._

```{r}
# check out the variables
# get_variables(modelFit)

modelFit %>%
    spread_draws(b_Intercept, sd_schoolid__Intercept, sigma) %>%
    filter(`.iteration` > 4000) %>%
    gather("Parameter", "Estimate", b_Intercept:sigma) %>%
    mutate(Chain=as.factor(`.chain`)) %>%
    ggplot(aes(x=`.iteration`, y=Estimate, color=Chain, group=Chain)) +
    geom_line(alpha=.5) +
    theme_classic() +
    facet_wrap(~Parameter, nrow=3, scales="free_y") +
    labs(x="Iteration")
```

## Question 3
_Summarize the posterior distribution you obtain in graphical and tabular form._

```{r}
modelFit %>%
    spread_draws(b_Intercept, sd_schoolid__Intercept, sigma) %>%
    gather("Parameter", "Estimate", b_Intercept:sigma) %>%
    group_by(Parameter) %>%
    median_qi(.width=.95) %>%
    select(Parameter:`.upper`) %>%
    rename(`2.5 CI`=`.lower`, `97.5 CI`=`.upper`) %>%
    knitr::kable()
```

```{r}
reDF <- modelFit %>%
  spread_draws(b_Intercept, r_schoolid[schoolid,]) %>%
  median_qi(school_mean = b_Intercept + r_schoolid, .width = c(.95))

reDF %>%
    arrange(school_mean) %>%
    mutate(id=1:n()) %>%
    ggplot(aes(y = id, x = school_mean)) +
    geom_pointintervalh(size=.7) +
    theme_classic() +
    geom_vline(linetype=2, xintercept=summary(modelFit)$fixed[1]) +
    theme(
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
    labs(y="", x="Estimate")
```

## Question 4  
_For which school is the difference between the non-Bayesian estimate and the Bayesian estimate of the group mean the largest?  Why do you think this school has the largest difference?_

```{r}
mleFit <- lmer(math ~ 1 + (1|schoolid), data=DF)

coef(mleFit)$schoolid %>%
    mutate(schoolid=as.numeric(row.names(.))) %>%
    rename(estMLE=`(Intercept)`) %>%
    left_join(reDF) %>%
    mutate(estDiff=abs(estMLE-school_mean)) %>%
    arrange(-estDiff)

DF %>% group_by(schoolid) %>% summarise(n=n()) %>% arrange(n)
```